### 2023-04-03 reachable-resilience.tex
up
details

\problemx{Downward reachability problem}
{A transition system $\mathscr{S}=(S,\rightarrow)$, $s \in S$ and a downward-closed set $D
\subseteq S$.}
{$\exists s' \in D ~ s \to^* s'$? \newline}

%\"Ozkan
\begin{proposition}[Proposition 1 in \cite{DBLP:conf/gg/Ozkan22}]\label{post*}
For finite-branching WSTS%with strong upward compatibility
, a basis of $\uparrow \post^*(s)$ is computable for every state $s$ iff the downward-reachability problem is decidable.
%i 			.e. given a state $s$ of a WSTS
% in the regarded class 
%with strong upward compatibility 
%and a decidable downward-closed set $D$, it can be decided whether $\exists s' \in D ~ s \to^* s'$. 
\end{proposition}

% \alain{rappeler l'idée de la preuve}
The idea behind the proof is the following. For deciding whether a downward-closed set $D$ is reachable from $s$, we check whether
$B_{\uparrow \post^*(s)} \cap D = \emptyset$, equivalent to $\post^*(s)\cap D = \emptyset$ by
Lemma~\ref{Lemma intersection}. For the converse direction, we compute the sequence of upward-closed
$U_n = \uparrow post^{\leq n}(s)$ until it becomes stationnary. 
Decidability of downward reachability leads to the decidability of the following stop condition:
asking whether $S \setminus U_n$ is reachable from $s$.

The characterization is used to show that Petri nets are $\post^*$-effective. Since the downward reachability problem is decidable for Petri-nets, we deduce the following:

\begin{proposition}
Let $P = (N,M,W)$ be a Petri net and $s \in M$ a configuration of $P$, then one can compute a basis of $\uparrow \post^*(s)$.
\end{proposition}

\begin{proof}
By Proposition~\ref{post*}, a basis of $\uparrow \post^*(s)$ is computable for every state $s$ iff the downward-reachability problem is decidable.
Let us now show that the downward-reachability problem is decidable for Petri nets.
Let us consider a downward closed set $D$ in a Petri net.
Let us consider $B$ a basis of the upward closed set $S \setminus D$.

$D$ is  defined by ... states where your number of counter is bounded by a constant (deduce this from ... having to be smaller than elements of the basis $B$), and others where it is unbounded. You have $m$ counters where it is bounded by $n$
then you have $m \times n$ possibilities for the values in these counters
you have unlimited possibilities for the values of the other counters,
but that is where the submarking problem comes in handy,
i.e. you ask the submarking problem for all of the $m \times n$ possibilities.

So downward-set reachability is decidable for PN.
\end{proof}

It is well-known that Petri nets are WSTS with strong upward compatibility (Thm.~6.1 in \cite{DBLP:journals/tcs/FinkelS01}). 
Since VASS are WSTS with strong upward compatibility and since there is an algorithm that computes a finite basis of  $\uparrow \post^*(s)$, \cite{DBLP:conf/gg/Ozkan22} deduced that bounded state-resilience is decidable for VASS.
Hence state-resilience is decidable for both Petri nets and VASS.
However, the hypothesis that $\uparrow \post^*$ is computable cannot be tested in the general WSTS framework. Moreover, we may show:

@@

\begin{proof}
Let us show that Reset VASS, that are WSTSs with strong upward compatibility, don't enjoy the property that $\uparrow \post^*$ is computable.
Suppose that one is able to compute a finite basis of $\uparrow \post^*$ for Reset VASS. Then, one would be able to decide whether $0$ is reachable 
%in counter $i$ 
by examining if there is %some vector $v$ in the basis such that $v(i)=0$
$0$ in the basis%
. But reachability of $0$  %in a counter $i$ 
is undecidable for Reset VASS. 
% sont WSTS with strong compatibility. et l'accessibilité est undecidable.
%	$\uparrow post^*$ n'est pas calculable pour les LCS non plus car $\uparrow post^*= \uparrow \downarrow post^*$ n'est pas calculable ???
\end{proof}

@@

% \mathieu{I moved the proof to the appendix since it's a slight adjustment but takes a lot of room.}
\begin{proof}
See Appendix~\ref{appendix} for the complete proof.
\end{proof}

Since the Reset Petri Net zero reachability problem is undecidable in Reset Petri-Nets, the reduction implies the following:

@@

Keeping the $\uparrow post^*$ effectiveness hypothesis but loosening the strong compatibility one still yields some decidability result for the general state-resilience problem.

@@

\begin{proof}[Sketch]
\mathieu{ (Formalism should be made more explicit so that the proof can be really pertinent)}
\textcolor{red}{
Let us consider a LCS where erasing a message take one step. Then the upward compatibility is not strong. 
For instance with a rule $\delta(q,b) = \text{ add some }b $,
we have $(q,b) \to (q,bb)$, $(q, baaa) \geq (q,b)$ and
$(q, baaa) \to^* (q,bb)$ however it takes more than a single step. 
In this case then $\pred(\uparrow \{(q,b)\})$ is not upward-closed. Hence 
$post^* (s) \cap D \subseteq \pred(\uparrow \{(q,b)\})$
do not imply
$\uparrow post^* (s) \cap D \subseteq \pred(\uparrow \{(q,b)\})$ for a downward-closed set $D$.
In particular by lossiness $(q, \epsilon)$ would belong to
$post^* (s)$,
hence $(q,a)$, $(q,aa)$, etc. would belong to $\uparrow post^* (s)$ although not necessarily to
$post^* (s)$. $(q,aaa)$ not being in $pred(\uparrow \{ (q,b)\})$ means the algorithm 
from \cite{DBLP:conf/gg/Ozkan22} would deduce that $1$-rechable-resilience does not hold in this example when $\Safe=\uparrow \{(q,b)\}$ and $\Bad =S$,
because $\uparrow post^* (s) \cap S \not\subseteq \pred(\uparrow \{(q,b)\})$,
despite having 
$post^* (s) \cap S \subseteq \pred(\uparrow \{(q,b)\})$.
}

@@

\mathieu{In case we have  effective basis of $post^*$ and upward compatibility (not strong) then we don't know the decidability status of k-state-resilience and bounded-state-resilience. Not necessarily a big deal, but should write a comment or two about this.}

\iffalse
\begin{proposition}
\textcolor{red}{
{\sc bounded resilience} is undecidable for WSTS with upward compatibility with an effective basis of $post^*$.
CONJECTURE
}
\end{proposition}

\begin{proof}
 \textcolor{red}{pas certain ! à faire Mathieu}
\end{proof}
\fi

@@

The authors from \cite{DBLP:conf/gg/Ozkan22,DBLP:journals/corr/abs-2108-00889}
have only been concerned with the case $\Safe = \uparrow \Safe$
and $\Bad =\downarrow \Bad$, 
\mathieu{pas tout à fait vrai: dans le technical report \cite{DBLP:conf/gg/Ozkan22} il considère aussi le cas où Bad et Safe sont tous les deux clos par le haut mais aussi le cas Safe clos par le bas Bad clos par le haut.}
 however as mentioned in Section~3
the property for instance of mutual exclusion is often modelised with $\Safe = \downarrow \Safe$.
Hence it makes sense to consider this symmetrical case as well. \\

\mathieu{rappeler résultats de \cite{DBLP:conf/gg/Ozkan22}}

\begin{definition}{marked WSTS} 
A {\em marked} WSTS is a tuple $(S, \leq, \to, INIT)$
where $(S, \leq, \to)$ is a WSTS and $INIT \subseteq S$. 
If $INIT$ is finite, we call it {\em fin-marked}.
\end{definition}

\begin{definition}
A marked WSTS is
\begin{itemize}
\item lossy if $\downarrow \post^* (INIT) = \post^* (INIT)$,
\item $\bot$-bounded (bottom-bounded) if there is a natural number 
$\ell \geq 0$ s.t. $s_B \in \post^{\leq \ell} (s) $ for every $ s \in S$ and every element 
$s_B$ of a basis of $S$ with $s \geq s_B$.
\end{itemize}
\end{definition}

\mathieu{}

\begin{theorem}{\cite{DBLP:conf/gg/Ozkan22}}
The bounded-state-resilience problem is decidable for \textcolor{red}{fin-marked} 
strongly compatible WSTS
 which are 
 {\em \textcolor{red}{lossy} } 
 and 
 {\em \textcolor{red}{$\bot$-bounded}  }
 if $\Bad = \uparrow \Bad$ and $\Safe = \downarrow \Safe$.
\end{theorem}

@@

\begin{theorem}
The state resilience problem is undecidable for effective WSTS with  strong  compatibility 
when
$\Safe=\downarrow \Safe$
and $\Bad=\uparrow \Bad$.
\end{theorem}

\begin{proof}
Consider Reset VASS
in the case $\Safe$ contains only the minimal element %$0$ 
and $\Bad = S \setminus \Safe$, 
it follows the state resilience problem is equivalent 
to the problem of whether $0$ can be visited infinitely often. 
%
% Since reachability of $0$ is undecidable for Reset VASS, we conclude.
Hence the undecidability.
\end{proof}

\mathieu{But what if we are in the case of $\uparrow \post^*$ effectiveness?}

\mathieu{Écrire que c'est globalement symmétrique au cas d'avant - je pense pas nécessairement qu'on arrive à faire beaucoup mieux que de juste dire ça.}

@@

\subsection{Synthesis}

\iffalse
\mathieu{Should try to put all results in a single table}

@@

Results in the case $\Safe = \uparrow \Safe$ and $\Bad = \downarrow \Bad$:

\begin{center}
\begin{tabular}{ | l | c | c | c | c |}
\hline  Hypothesis & strong compatibility ~ & $\uparrow post^*$ effective & both  \\ \hline
   SRP & Undecidable (Thm~\ref{srp up down}) & Decidable (Thm~\ref{post srp})  & Decidable (Thm~\ref{postcomputable})\\ \hline
   BSRP & Undecidable (Corollary~\ref{bsrp up down}) &  ??  & Decidable (Thm~\ref{ref ozkan}) \\ \hline
      kSRP & Undecidable (Corollary~\ref{bsrp up down}) & ?? & Decidable (Thm~\ref{ref ozkan}) \\ \hline
 \end{tabular}
\end{center}

Results in the case $\Safe = \downarrow \Safe$ and $\Bad = \uparrow \Bad$:

>>>>

\problemx{Downward-reachability problem}
{A transition system $\mathscr{S}=(S,\rightarrow)$, $s \in S$ and a downward-closed set $D
\subseteq S$.}
{$\exists s' \in D ~ s \to^* s'$? \newline}

%\"Ozkan
\begin{proposition}[Proposition 1 in \cite{DBLP:conf/gg/Ozkan22}]
For finite-branching WSTS%with strong upward compatibility
, a basis of $\uparrow \post^*(s)$ is computable for every state $s$ iff the downward-reachability problem is decidable.
%i 			.e. given a state $s$ of a WSTS
% in the regarded class 
%with strong upward compatibility 
%and a decidable downward-closed set $D$, it can be decided whether $\exists s' \in D ~ s \to^* s'$. 
\end{proposition}

% \alain{rappeler l'idée de la preuve}
The idea behind the proof is the following. For deciding whether a downward-closed set $D$ is reachable from $s$, we check whether
$B_{\uparrow \post^*(s)} \cap D = \emptyset$, that is equivalent to $\post^*(s)\cap D = \emptyset$ by
Lemma~\ref{Lemma intersection}. For the converse direction, we compute the sequence of upward-closed sets
$U_n = \uparrow post^{\leq n}(s)$ until it becomes stationnary. 
Decidability of downward reachability leads to the decidability of the following stop condition:
asking whether $S \setminus U_n$ is reachable from $s$.

The characterization is used to show that Petri nets are $\post^*$-effective. It is well-known that Petri nets are WSTS with strong upward compatibility (Thm.~6.1 in \cite{DBLP:journals/tcs/FinkelS01}). 
Since VASS are WSTS with strong upward compatibility and since there is an algorithm that computes a finite basis of  $\uparrow \post^*(S)$, \cite{DBLP:conf/gg/Ozkan22} deduced that bounded state-resilience is decidable for VASS.
Hence state-resilience is decidable for Petri nets and VASS.
However, the hypothesis that $\uparrow \post^*$ is computable cannot be tested in the general WSTS framework. Moreover, we may show:

@@

\begin{proof}
Let us show that Reset VASS, that are effective WSTSs with strong upward compatibility, don't enjoy the property that $\uparrow \post^*$ is computable.
Suppose that one are able to compute a finite basis of $\uparrow \post^*$ for Reset VASS. Then, one would be able to decide whether an element $m \in Min(\uparrow \post^*)$ is reachable.
%in counter $i$ 
by examining if there is %some vector $v$ in the basis such that $v(i)=0$
$0$ in the basis%
. But reachability of $0$  %in a counter $i$ 
is undecidable for Reset VASS. 
% sont WSTS with strong compatibility. et l'accessibilité est undecidable.
%	$\uparrow post^*$ n'est pas calculable pour les LCS non plus car $\uparrow post^*= \uparrow \downarrow post^*$ n'est pas calculable ???
\end{proof}

@@

\mathieu{I moved the proof to the appendix since it's a slight adjustment but takes a lot of room.}

Since reachability of $0$ is undecidable in Reset Petri-Nets, the reduction implies the following:

@@

Keeping the $\uparrow post^*$ effectiveness hypothesis but loosening the strong compatibility one still yields some decidability result for the general reachable-resilience problem.

@@

\begin{proof}[Sketch]
Let us consider a LCS where erasing a message take one step. Then the upward compatibility is not strong. 
For instance with a rule $\delta(q,b) = \text{ add some }b $,
we have $(q,b) \to (q,bb)$, $(q, baaa) \geq (q,b)$ and
$(q, baaa) \to^* (q,bb)$ however it takes more than a single step.

In this case then $\pred(\uparrow \{(q,b)\})$ is not upward-closed. Hence 
$post^* (s) \cap D \subseteq \pred(\uparrow \{(q,b)\})$
do not imply
$\uparrow post^* (s) \cap D \subseteq \pred(\uparrow \{(q,b)\})$ for a downward-closed set $D$.
In particular by lossiness $(q, \epsilon)$ would belong to
$post^* (s)$,
hence $(q,a)$, $(q,aa)$, etc. would belong to $\uparrow post^* (s)$ although not necessarily to
$post^* (s)$. $(q,aaa)$ not being in $pred(\uparrow \{ (q,b)\})$ means the algorithm 
from \cite{DBLP:conf/gg/Ozkan22} would deduce that $1$-rechable-resilience does not hold in this example when $\Safe=\uparrow \{(q,b)\}$ and $\Bad =S$,
because $\uparrow post^* (s) \cap S \not\subseteq \pred(\uparrow \{(q,b)\})$,
despite having 
$post^* (s) \cap S \subseteq \pred(\uparrow \{(q,b)\})$.

@@

\begin{proposition}
\textcolor{red}{
{\sc bounded resilience} is undecidable for WSTS with upward compatibility with an effective basis of $post^*$.
CONJECTURE
}
\end{proposition}

\begin{proof}
 \textcolor{red}{pas certain ! à faire Mathieu}
\end{proof}

@@

The authors from \cite{DBLP:conf/gg/Ozkan22,DBLP:journals/corr/abs-2108-00889}
have only been concerned with the case $\Safe = \uparrow \Safe$
and $\Bad =\downarrow \Bad$, however as mentioned in Section~3
the property for instance of mutual exclusion is often modelised with $\Safe = \downarrow \Safe$.
Hence it makes sense to consider this symmetrical case as well. \\

@@

\begin{theorem}
The resilience problem is undecidable for effective WSTS with  strong  compatibility 
when
$\Safe=\downarrow \Safe$
and $\Bad=\uparrow \Bad$.
\end{theorem}

\begin{proof}
Taking $\Safe = \{ 0 \}$, $\Bad = S \setminus 0$, it follows the state resilience problem is equivalent
to reachability of $0$. Since reachability of $0$ is undecidable for Reset VASS, we conclude.
\end{proof}

But what if we are in the case of $\uparrow \post^*$ effectiveness?

LOREM IPSUM \\

@@

\subsection{Other}

\mathieu{Should try to put all results in a single table}

@@

\mathieu{Everything below should be moved to the applications section no?}

% Let us deduce that effective basis of $post^*$ ex reset PN ?

 \alain{resilience decidable for LCS ? LCS sont-ils strong (quelle sémantique ?) et $\uparrow post^*(S)$ is computable ?}

We may consider context-free grammars.

\begin{corollary}
{\sc resilience} is decidable for finite automata and context-free grammars (seen as transition systems)
\end{corollary}

\begin{proof}

  \textcolor{red}{à faire Mathieu}
Context-free grammars are WSTS \cite{DBLP:journals/tcs/FinkelS01} \alain{monotony ne semble pas strong...}avec $\uparrow post^*$ calculable (because the reachability set of a pushdown machine is an effectively computable regular language) ???? resilience decidable.
\end{proof}

\cite{DBLP:journals/ipl/BouajjaniEFMRWW00}
